---
title: "Data Visualization using Cancer Patient Data"
author: "Merve Ozdemir"
date: "2025-07-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
In this project, I practiced my data analysis and visualization skills with R, using a public dataset of cancer patients in Denmark. I examined demographic trends, cancer types, and age distributions to understand patterns in diagnosis and treatment.

I downloaded the "Cancer Patient Data in Denmark" dataset through Kaggle, a data science platform that contains publicly available datasets. The dataset contains 3,000 rows and 25 variables including cancer type, age, treatment details, risk factors, and survival outcomes.

Below, I share my code, plots and thought process throughout the project, as I explore the dataset and try to come up with interesting questions and hypotheses. 

## Getting started: Loading libraries and data

```{r}
# Load required packages
library(tidyverse)

# Read the dataset (assumes file is in the same directory)
data <- read_csv("cancer_data_denmark.csv")

# View first few rows
head(data)
```

## Exploring the data with basic plots

The first thing I was interested in is the cancer types included in the dataset. I thus started with a bar plot showing the frequency of different cancer types. When I first generated this plot I noticed that the difference in the count of different cancer types was very small compared to the total observations, so I used coord_cartesian() to "zoom in" on the y axis to better visualize the difference.

```{r}
# Bar plot
plot_cancer_type <- ggplot(data, aes(x = Cancer_Type)) +
  geom_bar(fill = "skyblue") +
  coord_cartesian(ylim = c(425, 525)) +
  labs(
    title = "Cancer Type Distribution",
    x = "Cancer Type",
    y = "Count"
  ) +
  theme_minimal()

ggsave("plots/cancer_type_distribution.png", plot = plot_cancer_type, width = 8, height = 6, dpi = 300, bg = "white")

plot_cancer_type
```

Next, I visualized the distribution of patient ages using a density plot. I added transparency (alpha) to make the background and gridlines more visible. 

```{r}
plot_age <- ggplot(data, aes(x = Age)) +
  geom_density(fill = "skyblue", color = "darkblue", alpha = 0.4) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
  labs(
    title = "Age Distribution of Cancer Patients",
    x = "Age",
    y = "Density"
  ) +
  theme_minimal()

ggsave("plots/age_distribution.png", plot = plot_age, width = 8, height = 6, dpi = 300, bg = "white")

plot_age
```

The distributions were similar to what I expected. Now that I have this most basic info about the patients in the dataset (i.e. cancer type and age), I want to combine this information to see whether there are any interesting trends I could observe. For instance, is there any cancer type where the patients diagnosed are overall "younger" than in others? 

I generated a violin plot to see the age distribution across cancer types in my dataset. 

```{r}
plot_age_vs_cancertype <- ggplot(data, aes(x = Cancer_Type, y = Age)) + geom_violin(fill = "skyblue", alpha = 0.4) + labs(title = "Age Distribution by Cancer Type",
       x = "Cancer Type",
       y = "Age") +
  theme_minimal() 

ggsave("plots/age_vs_cancertype.png", plot = plot_age_vs_cancertype, width = 8, height = 6, dpi = 300, bg = "white")

plot_age_vs_cancertype
```

I wondered if age also factors into survival of patients. 

```{r}
plot_age_vs_survival <- ggplot(data, aes(x = Age, y = Survival_Months)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_smooth(method = "loess", se = TRUE, color = "darkred") +  # trend line
  labs(title = "Survival Months by Age",
       x = "Age",
       y = "Survival Months") +
  theme_minimal()

ggsave("plots/age_vs_survival.png", plot = plot_age_vs_survival, width = 8, height = 6, dpi = 300, bg = "white")


plot_age_vs_survival
```

Interesting! I would have expected age to affect survival outcome, but that doesn't seem to be the case, at least not at first glance. Overall, there doesn't seem to be a starking difference in the age distribution across cancers. What about gender?

```{r}
plot_gender_vs_cancertype <- ggplot(data, aes(x = Cancer_Type, fill = Gender)) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Gender within Each Cancer Type",
       x = "Cancer Type",
       y = "Proportion",
       fill = "Gender") +
  theme_minimal()

ggsave("plots/gender_vs_cancertype.png", plot = plot_gender_vs_cancertype, width = 8, height = 6, dpi = 300, bg = "white")

plot_gender_vs_cancertype
```

Seems like gender differences, too, are quite subtle in all types of cancer included in the dataset. Next, I wanted to see whether there were differences in survival outcome across different cancers, whether some cancers were more "deadly" than others. Of course, survival depends on many factors but it is still interesting to look for overall trends.

```{r}
plot_survival_vs_cancertype <- ggplot(data, aes(x = Cancer_Type, y = Survival_Months)) +
  geom_boxplot(fill = "turquoise") +
  labs(title = "Survival Months by Cancer Type",
       x = "Cancer Type",
       y = "Survival Months") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("plots/survival_vs_cancertype.png", plot = plot_survival_vs_cancertype, width = 8, height = 6, dpi = 300, bg = "white")

plot_survival_vs_cancertype
```

Similarly, do certain cancers have more occurances of relapse?


```{r}
plot_relapse_vs_cancertype <- ggplot(data, aes(x = Cancer_Type, fill = Relapse)) + scale_y_continuous(labels = scales::percent_format()) + geom_bar(position = "fill") + labs(
    title = "Relapse Rates by Cancer Type",
    x = "Cancer Type",
    y = "Percentage of Patients",
    fill = "Relapse"
  ) +
  theme_minimal()

ggsave("plots/relapse_vs_cancertype.png", plot = plot_relapse_vs_cancertype, width = 8, height = 6, dpi = 300, bg = "white")

plot_relapse_vs_cancertype

```

So far the data seems quite well balanced on these fronts! I next looked at patient habits such as smoking and alcohol consumption, which are known to cause cancer and therefore could point to certain trends... First, a general look at these habits without distinguishing between different cancer types:

```{r}
plot_smoking <- ggplot(data, aes(x = Cancer_Type, fill = Smoking_Status)) +
  geom_bar(position = "dodge") +  # proportions within cancer type
  labs(title = "Smoking Status Distribution by Cancer Type",
       x = "Cancer Type",
       y = "Count",
       fill = "Smoking Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("plots/smoking.png", plot = plot_smoking, width = 8, height = 6, dpi = 300, bg = "white")

plot_smoking

```

At the first glance, what was most interesting to me with this plot is that there are more non-smokers who were diagnosed with lung cancer than former/current smokers. But of course, smokers in this dataset are separated into "former" and "current" smokers. Unfortunately, I don't have enough information about the "study" conducted (since it isn't real--this dataset was created for educational purposes) I don't know what "former" implies; these patients could have quit smoking years ago, or maybe only for a week. So I'm not so interested in this labeling, I am rather interested in comparing people who never smoked with people who are (or were, at one point) smokers. 

```{r}
lung_cancer_data <- data %>% filter(Cancer_Type == "Lung") %>% mutate(Smoking_Status = ifelse(Smoking_Status == "Never", "never", "smoker")) #reminder of syntax: ifelse(condition, value_if_true, value_if_false)

plot_lung_cancer_data <- lung_cancer_data %>% ggplot(aes(x = Cancer_Type, fill = Smoking_Status)) + geom_bar(position = "dodge") + labs(title = "Smoking Status Distribution by Cancer Type",
       x = "Cancer Type",
       y = "Count",
       fill = "Smoking Status") +
  theme_minimal()

ggsave("plots/lung_cancer.png", plot = plot_lung_cancer_data, width = 8, height = 6, dpi = 300, bg = "white")

plot_lung_cancer_data
```

Now this makes a lot more sense to me visually! And what if we used these same simplified smoker vs. non-smoker categories to recreate the plot from before, across all cancer types?

```{r}
smokers_simplified <- data %>% mutate(Smoking_Status = ifelse(Smoking_Status == "Never", "never", "smoker"))
plot_smoking_simplified <- ggplot(smokers_simplified, aes(x = Cancer_Type, fill = Smoking_Status)) +
  geom_bar(position = "dodge") +  # proportions within cancer type
  labs(title = "Smoking Status Distribution by Cancer Type",
       x = "Cancer Type",
       y = "Count",
       fill = "Smoking Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("plots/smoking_simplified.png", plot = plot_smoking_simplified, width = 8, height = 6, dpi = 300, bg = "white")

plot_smoking_simplified
```

As expected, smokers (former and current) make up a higher proportion of patients in each cancer type. 

Before further analysis, I want to repeat these steps for alcohol consumption.


```{r}
plot_alcohol <- ggplot(data, aes(x = Cancer_Type, fill = Alcohol_Consumption)) +
  geom_bar(position = "dodge") +
  labs(title = "Alcohol Consumption by Cancer Type",
       x = "Cancer Type",
       y = "Count",
       fill = "Alcohol Consumption") +
  theme_minimal()

ggsave("plots/alcohol.png", plot = plot_alcohol, width = 8, height = 6, dpi = 300, bg = "white")

plot_alcohol

alcohol_simplified <- data %>% mutate(Alcohol_Consumption = ifelse(Alcohol_Consumption == "None", "none", "drinker"), Alcohol_Consumption = factor(Alcohol_Consumption, levels = c("none", "drinker"))
  )
plot_alcohol_simplified <- alcohol_simplified %>% ggplot(aes(x = Cancer_Type, fill = Alcohol_Consumption)) +
  geom_bar(position = "dodge") +
  labs(title = "Alcohol Consumption by Cancer Type",
       x = "Cancer Type",
       y = "Count",
       fill = "Alcohol Consumption") +
  theme_minimal()

ggsave("plots/alcohol_simplified.png", plot = plot_alcohol_simplified, width = 8, height = 6, dpi = 300, bg = "white")

plot_alcohol_simplified

```

I was also interested in the family history of these patients. This is a good way to test the "genetic" component of cancer, though of course there are always many factors at play. 

```{r}
plot_family_history <- ggplot(data, aes(x = Cancer_Type, fill = Family_History)) + geom_bar(position = "fill") + labs(title = "Family History across Cancer Types", x = "Cancer Type", y = "Proportion", fill = "Family History") + theme_minimal()

ggsave("plots/family_history.png", plot = plot_family_history, width = 8, height = 6, dpi = 300, bg = "white")

plot_family_history
```

Next, I wanted to see what kind of treatments these patients received, how common each treatment was, and whether one type of cancer got treated with a specific treatment more than others, etc. Below I explored the trends with different types of treatments included in the dataset. 

```{r}
data %>% count(Treatment_Type)

plot_treatment_facet <- data %>%
  ggplot(aes(x = Treatment_Type, fill = Treatment_Type)) +
  geom_bar(position = "dodge") +
  facet_wrap(~ Cancer_Type) + #facet_wrap creates a subplot for each cancer type
  coord_flip() + #flipped coordinates bc labels are easier to read horizontally
  labs(
    title = "Treatment Types by Cancer Type",
    x = NULL,  
    y = "Count",
    fill = "Treatment Type"
  ) +
  theme_minimal() 

ggsave("plots/treatments.png", plot = plot_treatment_facet, width = 8, height = 6, dpi = 300, bg = "white")

plot_treatment_facet


```

I noticed that there is also a variable called "Stage," which depicts which stage the cancer was diagnosed in. I wondered if there are any links between cancer progression and the type of treatment provided. 

```{r}
plot_treatment_per_stage <- ggplot(data, aes(x = Stage, fill = Treatment_Type)) + geom_bar(position = "fill") + labs(title = "Types of Treatment for Each Cancer Stage", x = "Stage", y = "Proportion") + theme_minimal()

ggsave("plots/treatment_per_stage.png", plot = plot_treatment_per_stage, width = 8, height = 6, dpi = 300, bg = "white")

plot_treatment_per_stage

plot_stage_per_treatment <- ggplot(data, aes(x = Treatment_Type, fill = Stage)) + geom_bar(position = "fill") + labs(title = "Types of Treatment for Each Cancer Stage", x = "Treatment", y = "Proportion") + theme_minimal()

ggsave("plots/stage_per_treatment.png", plot = plot_stage_per_treatment, width = 8, height = 6, dpi = 300, bg = "white")

plot_stage_per_treatment
```

It seems like every therapy has been used with cancers at every stage in this dataset. That is good to know; the dataset once again seems to be very balanced.

Finally, I wanted to take a look at the mutations variable of the dataset.

```{r}
count(data, Genetic_Mutation)
```
Seems like there are only four observations for genetic mutation: BRCA1, BRCA2, TP53, and none. 

```{r}
plot_mutations <- ggplot(data, aes(x = Cancer_Type, fill = Genetic_Mutation)) + geom_bar(position = "dodge") + labs(title = "Genetic Mutations across Cancer Types", x = "Cancer Type", fill = "Genetic Mutation") + theme_minimal()

ggsave("plots/mutations.png", plot = plot_mutations, width = 8, height = 6, dpi = 300, bg = "white")

plot_mutations
```

```{r}
mutations <- data %>% filter(Genetic_Mutation %in% c("BRCA1", "BRCA2", "TP53"))
mutations

plot_cancertype_with_mutation <- mutations %>% ggplot(aes(x = Cancer_Type)) + geom_bar(fill = "skyblue", width = 0.5) + coord_cartesian(ylim = c(200, 400)) + labs(title = "Types of Cancers with mutations identified", x = "Cancer Type", y = "Count") + theme_minimal()

ggsave("plots/cancertype_mutations.png", plot = plot_cancertype_with_mutation, width = 8, height = 6, dpi = 300, bg = "white")

plot_cancertype_with_mutation
```

Seems like lung cancer patients of this dataset have the most number of mutations identified. From one of the earlier plots we know that all three mutations "BRCA1", "BRCA2" and "TP53" are observed at similar counts in this cancer type, so I didn't plot that again separately. Instead, I decided to "zoom in" on this particular cancer type. To practice, I decided to try to build a predictive model of survival for lung cancer patients, using the data I have at hand. 



## Building a predictive model of survival for lung cancer

### First steps & setting up the training data

First, I need to set the stage for what I want to do. Below, I documented some preliminary steps, some of which were completely new to me, so I tried to take good notes.

```{r}
#First, filter the data to get a data frame with only lung cancer cases 
lung_data <- data %>%
  filter(Cancer_Type == "Lung") %>%
  drop_na(Survival_Months)

#Next, select the columns/variables you want to keep as "predictors" of survival
lung_model_data <- lung_data %>% 
  select(Survival_Months, Age, Stage, Smoking_Status, Alcohol_Consumption)

#training
set.seed(123) #makes sure random split is repeatable, otherwise i'd get a new split each time
train_indices <- sample(nrow(lung_model_data), 0.8 * nrow(lung_model_data)) #This gives me a random list of row numbers to be used for "training," all of which are saved in train_indices
#sample() randomly picks 80% of rows 
#nrow() total number of rows in the dataset

train_data <- lung_model_data[train_indices, ]
test_data <- lung_model_data[-train_indices, ] #all the rows that are remaining (bc of the -)
```

In these steps I filtered and selected certain factors from my dataset in order to create the data frame I'll be working with, which consists only of lung cancer patients and the "predictor" variables I decided to focus on. In the last section I included a "training/testing" step, where I "split" the data into two: one part (80%) used for training the model, the rest (20%) used to test the model. 

I learned that if you use the exact same data to both "train" and "test" your model, even if it's not working properly (i.e. if it is only memorizing things rather than actually "learning") everything will seem to work perfectly. To make sure that this isn't happening, I train and test the model with different parts of my data. Basically, I make sure that the model works even with new, previously unseen data.

Now, it's time to build the model!

### Building the linear regression model

During the building of my model, I encountered many unexpected problems, which were not errors in the code but rather outputs that just didn't "make sense," which led me to realize that I had overlooked some flaws in my data.

For instance, I learned that modeling is different for numeric/continuous variables (such as age) and categorial variables (such as smoking status); with a numeric/continuous variable like age, the model estimates a coefficient to represent the change in survival in the case of a one-unit (!) change in the variable (e.g. one additional year of age, one extra month, etc). But in the case of a categorical variable, it requires a "baseline" to compare every other category to. This baseline acts as a sort of reference. 

When I first set up my model, I realized that the output was only showing non-smokers and former-smokers, current-smokers were not present because the model had used this data as the "reference/baseline," which, of course, doesn't make much sense. If anything, the baseline for this variable should have been non-smokers. I learned that R uses whatever "level" is at the top as the baseline in such a case, so if I wanted the reference variable to be accurate, I needed to reorder the layers. 

To do this, I needed to convert all my character variables to factors, and then re-level. Below, I started by doing this simple reorganization, followed by the building of the final model.

```{r}
# Convert character columns to factors
train_data <- train_data %>%
  mutate(
    Stage = as.factor(Stage),
    Smoking_Status = as.factor(Smoking_Status),
    Alcohol_Consumption = as.factor(Alcohol_Consumption)
  )

# Re-level to set up the correct baseline
train_data$Smoking_Status <- relevel(train_data$Smoking_Status, ref = "Never")
train_data$Alcohol_Consumption <- relevel(train_data$Alcohol_Consumption, ref = "None")


model <- lm(Survival_Months ~ Age + Stage + Smoking_Status + Alcohol_Consumption, data = train_data)

summary(model)
```

Great! Now I had built my linear regression model to predict Survival_Months. I spent some time going over the elements of the output in order to understand exactly what I was seeing. Here, I document some of my key findings:

The "intercept" is the starting point of the prediction, showing me the predicted survival when all of the other variables are at their baseline value. The "coefficients," on the other hand, consider how the predicted survival would change (!) if one of the variables would increase or decrease, assuming that all others remain at baseline. This way, there are estimates for the influence of each variable individually. 

For example, the estimate value -0.023 for "Age" means that every additional year of age decreases survival months by this much, obviously a very tiny effect, also identified as being statistically insignificant by our model. This makes sense, considering I already plotted the age vs. survival data and saw that age didn't drastically influence survival months of a patient.

The statistically significant estimates of the model are usually marked by an asterisks (*), and since I see none here, I concluded that despite some differences, no variable has an effect on survival beyond whatever is possible by chance. 
 
Finally, I took a look at the Multiple R-squared: 0.02456 value, which let me know that only 3.2% of the variation in survival is explained by this model. Similarly, the p-value of 0.26 suggests overall model is not statistically significant.


## Conclusions 

Overall, this project was great for me to practice data visualization with R, and to try out building a linear regression model for the first time. There is still a lot I have to learn, especially in regards to modeling, so I am perfectly aware that this was not in any way perfect. What matters to me at the moment is to actively try to improve my skills! 

My main conclusion was that this dataset was very balanced; I failed to identify any variables that showed a statistically significant correlation with either survival outcome or cancer type. In all the cases I explored, there was an almost equal division between the different values of the variables. I am curious how it would be like working with a real dataset rather than a synthetic one, and will maybe try that next time! 

My model using lm() also confirmed my findings from the data viz, which was that there was no statistically significant "predictor" of lung cancer survival that I could use from this dataset using such a model. To put this finding in the context of the real world, lung cancer survival likely depends on more detailed clinical variables — like exact tumor size, genetic mutations, treatment response, etc. This dataset didn't have many clinical variables, and I also chose to focus on only a subset of variables for the sake of simplicity during my practice. The reality is much more nuanced, and many predictors may be weakly related or have overlapping effects. 


## Possible future work

I would definitely like to return to this project once I have improved my data viz and modeling skills. I would then add to my code and explore the dataset in newer, more advanced ways, maybe taking this one step further. For instance, here I only focused on linear regression since that is all I know so far, but I would like to practice using decision trees or random forests in the future, as these can handle non-linear patterns better than linear regression.

I would also like to try modeling using more clinical predictors, such as tumor size, metastasis, lymph node examined, etc. 

I coded the entire project in R, but I also enjoy coding in Python, and would consider using the same dataset for analysis in Python, just to see the differences in the process! 


## Feedback

As always, I am open to any feedback! I am a wet-lab biologist who is completely self-taught in anything computer-related, and I still have a long way to go to reach the level I want to be at! If you noticed any errors in my code/thinking, or have suggestions on how I could improve this project, reach out! Any feedback from you is always welcome and appreciated. Thanks very much :-) 




