---
title: "Data Visualization using Cancer Patient Data"
author: "Merve Ozdemir"
date: "2025-07-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Introduction
This report explores synthetic cancer patient records from Denmark. We examine demographic trends, cancer types, and age distributions to understand patterns in diagnosis and treatment.

The dataset contains 3,000 rows and 25 variables including cancer type, age, treatment details, risk factors, and survival outcomes.

#Load Libraries and Data

```{r}
# Load required packages
library(tidyverse)

# Read the dataset (assumes file is in the same directory)
data <- read_csv("cancer_data_denmark.csv")

# View first few rows
head(data)
```

#Plot 1: Distribution of Cancer Types
We start with a bar plot showing the frequency of different cancer types.

```{r}
# Bar plot
plot_cancer_type <- ggplot(data, aes(x = Cancer_Type)) +
  geom_bar(fill = "skyblue") +
  coord_cartesian(ylim = c(425, 525)) +
  labs(
    title = "Cancer Type Distribution",
    x = "Cancer Type",
    y = "Count"
  ) +
  theme_minimal()

# Display the plot
plot_cancer_type
```
#Plot 2: Age Distribution of Patients
Next, we visualize the distribution of patient ages using a density plot. Transparency (alpha) is added to make the background and gridlines more visible.

```{r}
plot_age <- ggplot(data, aes(x = Age)) +
  geom_density(fill = "skyblue", color = "darkblue", alpha = 0.4) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
  labs(
    title = "Age Distribution of Cancer Patients",
    x = "Age",
    y = "Density"
  ) +
  theme_minimal()

plot_age
```

